Blessing of FedAvg's Linear Speedup in Federated Learning

### 2020-05-27 source/sec_linear_regression.tex
update figures
interpolation results and proofs updated

\section{Linear Regression}

@@

We first show that FedAvg with MaSS has exponential convergence for
linear regression when the global objective has 0 as its global minimum.

@@

We now present the exponential convergence result in linear regression
using FedAvg with MaSS updates. On each device, local data is stored
and mini-batch gradient descent with batch size $m$ is performed.
We assume that the batch size is the same across devices. 
\begin{theorem}
	(FedAvg with MaSS, Linear Regression) Let $\tilde{\kappa}_{m}:=\tilde{\kappa}/m+(m-1)/m$,
	and let the hyperparameters satisfy 
	\begin{align*}
	\eta^{k}(m)=\frac{1}{L_{m}}, & \alpha^{k}(m)=\frac{1}{\sqrt{\kappa_{m}\tilde{\kappa}_{m}}},\delta^{k}(m)=\frac{\eta^{k}}{\alpha^{k}\tilde{\kappa}_{m}}
	\end{align*}
	for all $k$. Let $\mathcal{I}_{E}=\{\ell E\mid\ell\in\mathbb{N}\}$
	be the set of communication rounds, then the FedAvg algorithm with
	MaSS and full participation solves the problem 
	\begin{align*}
	F(w) & =\sum_{k=1}^{N}p_{k}F_{k}(w)\\
	F_{k}(w) & =\frac{1}{2n_{k}}\sum_{j=1}^{n_{k}}(w^{T}x_{k,j}-z_{k,j})^{2}
	\end{align*}
	with exponential convergence
	\begin{align*}
	\mathbb{E}F(\overline{w}_{t})\leq\frac{L}{2}\mathbb{E}\|\overline{w}_{t}-w^{\ast}\|^{2} & \leq C\cdot\exp(-\frac{t}{\sqrt{(\kappa_{N}+(E-1)^{2})(\tilde{\kappa}_{N}+(E-1)^{2})}})
	\end{align*}
	where $w^{\ast}$ is such that $F(w^{\ast})=0$ and $C=\frac{L}{2}\cdot\frac{\alpha}{\delta}\sum_{k=1}^{N}p_{k}(\|v_{0}^{k}-w^{\ast}\|_{H_{k}^{-1}}^{2}+\frac{\delta}{\alpha}\|w_{0}^{k}-w^{\ast}\|^{2})$.

In particular, this implies that for $E=1$, there is linear speedup, i.e. 
	\begin{align*}
	\|\overline{w}_{t+1}-w^{\ast}\|^{2} & \leq C\cdot\exp(-\frac{t}{\sqrt{(\kappa_{N})(\tilde{\kappa}_{N})}})\leq C\cdot\exp(-\frac{Nt}{\sqrt{(\kappa_{1})(\tilde{\kappa})}})
	\end{align*}

	With partial participation of $K$ devices each communication round,
	the convergence rate is also 
	\begin{align*}
	\mathbb{E}F(\overline{w}_{t})\leq\frac{L}{2}\mathbb{E}\|\overline{w}_{t}-w^{\ast}\|^{2} & \leq C\cdot(1+\frac{4}{K}\cdot(\frac{t}{E}-1))\cdot(1-\frac{1}{\sqrt{\kappa_{K}\tilde{\kappa}_{K}}})^{t}=O(\exp(-\frac{t}{\sqrt{(\kappa_{N}+(E-1)^{2})(\tilde{\kappa}_{N}+(E-1)^{2})}}))
	\end{align*}

\end{theorem}

>>>>

\section{Exponential Convergence of FedAvg in the Overparameterized Setting}

@@

We first show that FedAvg with SGD has exponential convergence for
linear regression when the global objective has 0 as its global minimum.

\subsection{Exponential Convergence of FedAvg with SGD}

In Belkin et al., the authors show that in the interpolation setting,
SGD achieves exponential convergence with linear speedup batch size
that is below some critical threshold. In this section, we extend
their result to the federated learning setting and show that similar
exponential convergence with linear speedup in the number of workers
hold using FedAvg with SGD updates holds.

We consider the standard setting of empirical risk minimization where
$\ell(w;\xi_{k,j})$ is non-negative, $l$-smooth, and convex, and
as before, each $F_{k}(w)=\frac{1}{n_{k}}\sum_{j=1}^{n_{k}}\ell(w;\xi_{k,j})$
is $L$-smooth and $\mu$-strongly convex. Note that $l\geq L$. This
setting includes many important problems in practice, such as linear
regression with a full rank sample covariance matrix. In the interpolation
or overparameterized setting, we assume that there exists $w^{\ast}\in\arg\min_{w}\sum_{k=1}^{N}p_{k}F_{k}(w)$
such that $\ell(w^{\ast};\xi_{k,j})=0$ for all $\xi_{k,j}$. We first
present a general result on the exponential convergence with linear
speedup in the number of workers, before giving a sharper analysis
for linear regression problems. 
\begin{theorem}
	For the overparameterized setting with general strongly convex and
	smooth objectives, FedAvg with local SGD updates and communication
	every $E$ iterations with constant step size $\overline{\alpha}=\frac{1}{2E}\frac{N}{l\nu_{\max}+L(N-\nu_{\min})}$
	gives the exponential convergence guarantee 
	\begin{align*}
	\mathbb{E}F(\overline{w}_{t}) & \leq\frac{L}{2}(1-\mu\overline{\alpha})^{t}\|w_{0}-w^{\ast}\|^{2}=O(\exp(-\frac{\mu}{2E}\frac{N}{l\nu_{\max}+L(N-\nu_{\min})}t)\cdot\|w_{0}-w^{\ast}\|^{2})
	\end{align*}
\end{theorem}
%
We see that when $l>L$, the speedup factor is on the order of $\frac{N}{E(l\nu_{\max}+L(N-\nu_{\min}))}/\frac{1}{l\nu_{\max}+L(1-\nu_{\min})}\approx\frac{Nl}{E(l+L(N-1))}=O(N/E)$
for $N\leq\frac{l}{L}+1$, i.e. FedAvg with $N$ workers and communication
every $E$ iterations provide an exponential convergence speedup factor
of $O(N/E)$, for $N\leq\frac{l}{L}+1$.

Next we show that specialized to the overparameterized linear regression
setting, the exponential convergence bound can be further sharpened,
with a similar linear speedup in the number of workers. In this setting,
we have 
\begin{align*}
F(w) & =\sum_{k=1}^{N}p_{k}F_{k}(w)\\
F_{k}(w) & =\frac{1}{2n_{k}}\sum_{j=1}^{n_{k}}(w^{T}x_{k,j}-z_{k,j})^{2}
\end{align*}
and there exists $w^{\ast}$ such that $F_{k}(w^{\ast})\equiv0$
for all $k$. 
\begin{theorem}
	(FedAvg with SGD, Linear Regression) For the overparamterized linear
	regression problem, FedAvg with local SGD updates and communication
	every $E$ iterations with constant step size $\overline{\alpha}=\frac{1}{2E}\frac{N}{l\nu_{\max}+L(N-\nu_{\min})}$
	gives the exponential convergence guarantee 
	\begin{align*}
	\mathbb{E}F(\overline{w}_{t}) & \leq L(1-c\frac{N}{E\kappa})^{t}\|w_{0}-w^{\ast}\|^{2}
	\end{align*}
	where $\kappa=L/\mu$ and $c$ is some universal constant. 
\end{theorem}

Next we show FedAvg with MaSS has exponential convergence but with an improved rate.

@@

We now present the exponential convergence result in linear regression
using FedAvg with MaSS updates. On each device, local data is stored
and mini-batch gradient descent with batch size $m$ is performed.
We assume that the batch size is the same across devices. 
\begin{theorem}
	(FedAvg with MaSS, Linear Regression) For the overparamterized linear
	regression problem, FedAvg with local MaSS updates and communication
	every $E$ iterations with constant step size $\overline{\alpha}=\frac{1}{2E}\frac{N}{l\nu_{\max}+L(N-\nu_{\min})}$
	gives the exponential convergence guarantee 
	\begin{align*}
	\mathbb{E}F(\overline{w}_{t}) & \leq L(1-c\frac{N}{E\sqrt{\kappa\tilde{\kappa}}})^{t}\|w_{0}-w^{\ast}\|^{2}
	\end{align*}
	where $\kappa=L/\mu$, $c$ is some universal constant, and $\tilde{\kappa}$
	is the smallest positive number such that $\mathbb{E}\|\tilde{x}\|_{H^{-1}}^{2}\tilde{x}\tilde{x}^{T}\preceq\tilde{\kappa}H$. 
	\label{th:fedmasslr}
\end{theorem}

We immediately see that since $\tilde{\kappa}\leq\kappa$, FedAvg with MaSS provides an improvement over FedAvg with SGD in the the interpolation setting, while enjoying the same speedup factor in the number of workers.
